Random Forests

### 2020-07-02 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
small corrections in section 'aggregation'; added section on distributional forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3035 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c8e5510b3c8d99b30497ac311a8f39e92a900e30

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
%\usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usepackage{calc}
\usepackage{graphicx}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
\usepackage{xcolor}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

@@

library("partykit")
library("Formula")
library("latex2exp")
library("lattice")
library("MASS")
library("colorspace")

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

@@

\visible<13->{
for $\hat{y}_t$ being the average response value in the segment $\mathcal{B}^t_x$ which the observed covariate $x$ is assigned to in the $t$-th tree with learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:

@@

$\hat{y}_t= \frac{1}{|\mathcal{B}^t_x|} \sum_{i: y_i \in \mathcal{B}^t_x} y_i$
\visible<14->{
$\ =\ \sum_{i=1}^n \frac{I(x_i \in \mathcal{B}^t_x)}{|\mathcal{B}^t_x|} \cdot y_i$
}
\visible<15->{
$\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$
}
}
\end{frame}

@@

\begin{frame}
\frametitle{Aggregation of trees}
{\bf Simple tree model:} Obtain the predicted response $\hat{y}$ for an observed covariate~$x$ by averaging over observations in the same terminal node $\mathcal{B}_x$: 
$$
\hat{y}= \frac{1}{|\mathcal{B}_x|} \sum_{i: y_i \in \mathcal{B}_x} y_i
\visible<2->{
\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$$
}

{\bf Model-based tree:} Obtain the predicted model parameter(s) $\hat{\theta}$ for an observed covariate~$x$ by optimizing an objective function $f$ for all observations in the same terminal node $\mathcal{B}_x$:
$$
\hat{\theta}= \argmin_{\theta} \sum_{i: y_i \in \mathcal{B}_x} f(\theta; y_i)
\visible<2->{
\ =\ \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)$$
}

{\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (y_i - \theta)^2$.

@@

The predicted model parameter (vector) $\hat{\theta}$ 
is the argument of the parameter space $\Theta$ 
that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:

@@

\[
\hat{\theta}(x) =  \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)
\]

@@

{\bf Sampling observations:}
\begin{itemize}
\item Bootstrapping: resampling with replacement, usually ``n out of n''.
\item Subsampling: draw a subset without replacement (typically of the size of 63.2\% of the full data set, but can be a smaller fraction for large data sets).
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree I:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest1, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[1]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree II:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest2, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[2]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree III:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest3, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[3]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree IV:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest4, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[4]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

@@

  cf <- cforest(f, data = CPS1985, ntree = 4)
  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
  colnames(predwage) <- c("age", "predicted log(wage)")
  plot(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
       ylab = "predicted wage", xlab = "age", main = "Forest with 4 trees")

@@

{\bf Idea:} Asses the influence of a covariate on the forest model by evaluating the change in performance of the model (mean decrease in accuracy) after permutating the selected covariate.

@@

Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, number of misclassifications, \ldots).
%}
\end{frame}

@@

{\bf partykit:}
\begin{itemize}
  \item R package offering a toolkit for unbiased recursive partitioning. 
  \item Provides the forest-building function \fct{cforest}, based on conditional inference trees (Ctree).
\end{itemize}

@@

\textbf{Further R packages:}
%
\begin{itemize}
  \item grf: generalized random forests,
  \item disttree: distributional forests, based on partykit,
  \item \ldots
\end{itemize}

@@

>>>>

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
%\usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usepackage{calc}
\usepackage{graphicx}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
\usepackage{xcolor}
\usepackage{changepage}
%\usepackage[cal=boondoxo]{mathalfa}
%\graphicspath{{plots/}}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

@@

library("partykit")
library("Formula")
library("latex2exp")
library("lattice")
library("MASS")
library("colorspace")
library("disttree")
library("latex2exp")
library("gamlss")
library("crch")
library("RainTyrol")
library("ggplot2")
theme_set(theme_bw(base_size = 18))
library(colorspace)
## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

@@

\visible<14->{
for $\hat{y}_t$ being the average response value in the segment $\mathcal{B}^t_x$ which the observed covariate $x$ is assigned to in the $t$-th tree with learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:

@@

$\hat{y}_t= \frac{1}{|\mathcal{B}^t_x|} \sum_{i: y_i \in \mathcal{B}^t_x} y_i$
\visible<15->{
$\ =\ \sum_{i=1}^n \frac{I(x_i \in \mathcal{B}^t_x)}{|\mathcal{B}^t_x|} \cdot y_i$
}
\visible<16->{
$\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$
}
}
\end{frame}

@@

\begin{frame}
\frametitle{Aggregation of trees}
{\bf Simple tree model:} Obtain the predicted response $\hat{y}$ for an observed covariate~$x$ by averaging over observations in the same terminal node $\mathcal{B}_x$: 
$$
\hat{y}= \frac{1}{|\mathcal{B}_x|} \sum_{i: y_i \in \mathcal{B}_x} y_i
\visible<2->{
\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$$
}
\visible<3->{
$$\ =\ \argmin_{\mu} \sum_{i=1}^n w_i(x) \cdot (y_i - \mu)^2$$
}

\visible<4->{
{\bf Model-based tree:} Obtain the predicted model parameter(s) $\hat{\theta}$ for an observed covariate~$x$ by optimizing a loss function for all observations in the same terminal node $\mathcal{B}_x$:
$$
\hat{\theta}= \argmin_{\theta} \sum_{i: y_i \in \mathcal{B}_x} loss(\theta; y_i)
\ =\ \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot loss(\theta; y_i)$$
}

% {\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (y_i - \theta)^2$.

@@

The predicted model parameter (vector) $\hat{\theta}$ 
is the argument of the parameter space $\Theta$ 
that minimizes the weighted sum of a loss function evaluated for learning data $\{(y_i,x_i)\}_{i=1,\ldots,n}$:

@@

\[
\hat{\theta}(x) =  \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot loss(\theta; y_i)
\]

@@

{\bf Sampling observations:}
\begin{itemize}
\item Bootstrapping: resampling with replacement, usually ``n out of n''.
\item Subsampling: draw a subset without replacement (typically of the size of 63.2\% of the full data set; for large data sets a smaller fraction can be advantageous).
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree I:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest1, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[1]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree II:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest2, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[2]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree III:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest3, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[3]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree IV:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest4, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[4]], data = CPS1985[,c("wage", "education", "experience", "age", "ethnicity", "gender", "union")]), drop_terminal = TRUE)
@
\end{frame}

@@

  cf <- cforest(f, data = CPS1985, ntree = 4)
  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
  colnames(predwage) <- c("age", "predicted log(wage)")
  plot(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
       ylab = "predicted wage", xlab = "age", main = "Forest with 4 trees")
  lines(x = predwage[,"age"], y = exp(predwage[,"predicted log(wage)"]), type = 'l',
       col = "gray")

@@

{\bf Idea:} Asses the influence of a covariate on the forest model by 
\begin{itemize} 
\item permuting the selected covariate and evaluating the resulting change in performance of the model (mean decrease in accuracy, measured for example based on the employed loss function) or
\item calculating the total decrease in node impurities from splitting in the selected covariate (measured for example by the Gini index for classification trees or the RMSE for regression trees), 
\end{itemize}
averaged over all trees.

@@

% Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, number of misclassifications, \ldots).\\
% To measure impurity, for example, the Gini index can be employed for classification trees or the RMSE for regression trees.
%}
\end{frame}

@@

{\bf partykit:}
\begin{itemize}
  \item R package offering a toolkit for unbiased recursive partitioning. 
  \item Provides the forest-building function \fct{cforest}, based on conditional inference trees (CTree).
\end{itemize}

@@

\textbf{Further R packages:}
%
\begin{itemize}
  \item grf: generalized random forests,
  \item disttree: distributional forests, based on partykit,
  \item circtree: circular distributional forests, based on partykit,
  \item model4you: model-based random forests for personalized treatment effects,
  \item randomSurvivalForest: random forests for the analysis of right-censored survival data, 
  \item \ldots
\end{itemize}

@@

\subsection{Distributional forests}
\begin{frame}%[fragile]
\frametitle{Motivation}
\vspace{-0.41cm}
\begin{figure}[!htb]
\minipage{0.285\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<2->{
<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
\end{center}
\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.285\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<3->{
<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
lines(x = x, y = ytrue, col = pal["forest"], lwd = 7, main = "")
@
}
\end{center}
\endminipage
\visible<4->{{\LARGE$\rightarrow$}}
\minipage{0.285\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\visible<4->{
<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
}
\end{center}
\endminipage

\vspace{0.5cm}
\minipage{0.25\textwidth}
\begin{center}
\visible<2->{
LM, GLM\\
\vspace{0.5cm}
\code{lm}\\
\code{glm}\\
\vspace{1.5cm}}
\end{center}
\endminipage
\hspace{1.1cm}
\minipage{0.25\textwidth}
\begin{center}
\visible<3->{
GAM\\
\vspace{0.5cm}
\code{mgcv}\\
\code{VGAM}\\
\vspace{1.5cm}}
\end{center}
\endminipage
\hspace{1.1cm}
\minipage{0.25\textwidth}
\begin{center}
\visible<4->{
GAMLSS\\
\vspace{0.5cm}
\code{gamlss}\\
\code{mgcv}\\
\code{VGAM}\\
\code{gamboostLSS}\\
\code{bamlss}}
\end{center}
\endminipage
\end{figure}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
\vspace{-0.41cm}
\begin{figure}[!htb]
\minipage{0.285\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@
\visible<1->{
<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
}
\end{center}

\endminipage
\visible<2->{{\LARGE$\rightarrow$}}
\minipage{0.285\textwidth}
 \begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@
\visible<2->{
<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
}
\end{center}

\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.285\textwidth}
\begin{center}
\visible<3->{
<<plot_motivation_question, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
box(lwd=5)
text(x = mean(range(x)), y = mean(range(y)), "?", cex = 12)
@
}
\end{center}

\endminipage

\minipage{0.285\textwidth}
\begin{center}
\vspace{0.0cm}
\visible<1->{
Regression tree\\
\vspace{0.3cm}
\resizebox{0.2\textwidth}{!}{
\begin{tikzpicture}
  \node[ellipse, fill=HighlightBlue!70, align=center] (n0) at (1, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n1) at (0.5, 1) {};
  \draw[-, line width=1pt] (n0) -- (n1);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n2) at (1.5, 1) {};
  \draw[-, line width=1pt] (n0) -- (n2);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n3) at (1, 0) {};
  \draw[-, line width=1pt] (n2) -- (n3);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n4) at (2, 0) {};
  \draw[-, line width=1pt] (n2) -- (n4);
\end{tikzpicture}}
\vspace{0.2cm}\\
\code{rpart}\\
\code{party(kit)}\\
\vspace{1cm}}
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.285\textwidth}
\begin{center}
\vspace{-0.4cm}
\visible<2->{
Random forest\\
\vspace{0.4cm}
\resizebox{0.6\textwidth}{!}{
\begin{tikzpicture}
  \node[ellipse, fill=HighlightBlue!70, align=center] (n00) at (1, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n01) at (0.5, 1) {};
  \draw[-, line width=1pt] (n00) -- (n01);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n02) at (1.5, 1) {};
  \draw[-, line width=1pt] (n00) -- (n02);

  \node[ellipse, fill=HighlightBlue!70, align=center] (n10) at (3, 2) {};
  \node[ellipse, fill=HighlightBlue!70, align=center] (n11) at (2.5, 1) {};
  \draw[-, line width=1pt] (n10) -- (n11);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n12) at (3.5, 1) {};
  \draw[-, line width=1pt] (n10) -- (n12);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n13) at (2, 0) {};
  \draw[-, line width=1pt] (n11) -- (n13);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n14) at (2.8, 0) {};
  \draw[-, line width=1pt] (n11) -- (n14);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n15) at (3.2, 0) {};
  \draw[-, line width=1pt] (n12) -- (n15);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n16) at (4, 0) {};
  \draw[-, line width=1pt] (n12) -- (n16);

  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n21);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n22);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
  \draw[-, line width=1pt] (n22) -- (n23);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
  \draw[-, line width=1pt] (n22) -- (n24);
\end{tikzpicture}
}
\vspace{0.3cm}\\
\code{randomForest}\\
\code{ranger}\\
\code{party(kit)}}
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.285\textwidth}
\begin{center}
\vspace{0.15cm}
\visible<3->{
Distributional trees and forests\\
\vspace{1.15cm}
\code{disttree}\\
based on \code{partykit}\\
\vspace{1cm}}
\end{center}
\endminipage
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Building blocks}
%\vspace{0.5cm}

\begin{minipage}{0.79\textwidth}
\vspace{-0.5cm}
\textbf{Distributional modeling:}
\begin{itemize}
\vspace{-0.05cm}
  \item Specify the complete probability distribution \\
  \vspace{-0.05cm}
  (location, scale, shape, \dots).
\end{itemize}

\smallskip

\textbf{Tree:}
\begin{itemize}
\vspace{-0.05cm}
  \item Capture non-linear and non-additive effects.
  \vspace{-0.1cm}
  \item Automatic selection of covariates and interactions. %detection of steps and abrupt changes. %(data driven)
\end{itemize}

\smallskip

\textbf{Forest:}
\begin{itemize}
\vspace{-0.05cm}
  \item Smoother effects.
  \vspace{-0.1cm}
  \item Stabilization and regularization of the model.
\end{itemize}

$\Rightarrow$ \textbf{Synthesis:} Distributional Trees and Forests
\end{minipage}
<<motivation_GAMLSS_syn, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
nobs <- 200
x_g <- c(1:nobs)/nobs
x_g <- 2*(x_g-0.5)
ytrue_g <- x_g^3
var_g <- exp(-(1.74*x_g)^2)/2.5
y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var_g)
@
<<motivation_regtree_syn, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- var <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
for(i in 1:nobs) var[i] <-  if(x[i]<1/3) 0.1 else 0.6 *(4/3 - x[i])
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
@
<<motivation_randforest_syn, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@
\begin{minipage}{0.2\textwidth}
\vspace{-0.05cm}
<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g)),
  col = pallight["forest"], border = "transparent")
lines(x_g, ytrue_g, col = pal["forest"], lwd=10)
points(x_g, y_g, col = "slategray", pch = 19)
box(lwd = 5)
@

\vspace{0.15cm}

<<plot_motivation_regtree_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
     ylim = c(min(y), max(y)))
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=10)
box(lwd=5)
@

\vspace{0.15cm}

<<plot_motivation_randforest4_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
     ylim = c(min(y), max(y)))
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=10, main = "")
box(lwd=5)
@

\vspace{0.15cm}

<<plot_motivation_distforest, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
#plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
#box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
#lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19,
     ylim = c(min(y), max(y)))
polygon(c(x, rev(x)), c(yforest + 0.1 + var, rev(yforest - (0.1 + var))),
       # c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g))
  col = pallight["forest"], border = "transparent")
box(lwd=5)
lines(x = x, y = yforest, col = pal["forest"], lwd=10)
@
\end{minipage}
\end{frame}

<<treedata, echo=FALSE, results=hide>>=
set.seed(54)
nobs <- 500
x <- runif(nobs, 0, 1)
mu <- sigma <- ytrue <- numeric(length = nobs)
for(i in 1:nobs) sigma[i] <- if(x[i]<=0.4) 1 else 3
for(i in 1:nobs) mu[i] <- if(x[i]<= 0.4|| x[i]>0.8) 4 else 12
y <- rnorm(nobs, mean = mu, sd = sigma)
ytrue <- mu
data <- data.frame(cbind(y,x, ytrue))

alldata <- cbind(data, mu, sigma)
odata <- alldata[order(alldata["x"]),]
@

<<dgp_tree, echo=FALSE, results=hide>>=
data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
names(data) <- c("x","x","x")
fig <- party(
  partynode(1L,
            split = partysplit(2L, breaks = 0.4),
            kids = list(
              partynode(2L, info = c(
                "n = 200",
                "   True parameters:   ",
                expression(mu == '4'),
                expression(sigma == '1')
              )),
              partynode(3L,
                        split = partysplit(3L, breaks = 0.8),
                        kids = list(
                          partynode(4L, info = c(
                            "n = 200",
                            "   True parameters:   ",
                            expression(mu == '12'),
                            expression(sigma == '3')
                          )),
                          partynode(5L, info = c(
                            "n = 100",
                            "   True parameters:   ",
                            expression(mu == '4'),
                            expression(sigma == '3')
                          )))))),
  data
)

node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
    gp = gpar()) 
{
    meta <- obj$data
    nam <- names(obj)
    extract_label <- function(node) {
        if (is.terminal(node)) 
            return(rep.int("", 2L))
        varlab <- character_split(split_node(node), meta)$name
        if (abbreviate > 0L) 
            varlab <- abbreviate(varlab, as.integer(abbreviate))
        if (pval) {
            nullna <- function(x) is.null(x) || is.na(x)
            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
                silent = TRUE))
            pval <- if (inherits(pval, "try-error")) 
                FALSE
            else pval
        }
        if (pval) {
            pvalue <- node$info$p.value
            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
                paste("p =", round(pvalue, digits = 3L)))
        }
        else {
            plab <- ""
        }
        return(c(varlab, plab))
    }
    maxstr <- function(node) {
        lab <- extract_label(node)
        klab <- if (is.terminal(node)) 
            ""
        else unlist(lapply(kids_node(node), maxstr))
        lab <- c(lab, klab)
        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
        lab <- lab[which.max(nchar(lab))]
        if (length(lab) < 1L) 
            lab <- ""
        return(lab)
    }
    nstr <- maxstr(node_party(obj))
    if (nchar(nstr) < 6) 
        nstr <- "aAAAAa"
    rval <- function(node) {
        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
            height = unit(3, "lines"), name = paste("node_inner", 
                id_node(node), sep = ""), gp = gp)
        pushViewport(node_vp)
        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
            seq(0.8, 1, by = 0.01))
        yell <- sqrt(xell * (1 - xell))
        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
        lab <- extract_label(node)
        fill <- rep(fill, length.out = 2L)
        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
            "lines"))
        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
        #    "lines"))
        grid.text(lab[2L], y = unit(1, "lines"))
        if (id) {
            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
                "npc"), width = max(unit(1, "lines"), unit(1.3, 
                "strwidth", nam[id_node(node)])), height = max(unit(1, 
                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
            pushViewport(nodeIDvp)
            grid.rect(gp = gpar(fill = fill[2]))
            grid.text(nam[id_node(node)])
            popViewport()
        }
        upViewport()
    }
    return(rval)
}

class(node_inner_ext) <- "grapcon_generator"

\begin{frame}[fragile]
\frametitle{Distributional trees}
\vspace*{-0.12cm}
\begin{center} 
DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$

\vspace*{-0.21cm}
\setkeys{Gin}{width=0.58\linewidth}
<<plottree_xyplot, fig=TRUE, echo=FALSE,width=7>>=
par(mar=c(5.1,4.1,2.4,1.1))
plot(y=odata$y, x=odata$x, ylab = "y", xlab = "x", col = "gray")
lines(x = odata$x, y = odata$mu, col = pal["forest"], lwd = 2.5, main = "")
polygon(c(odata$x, rev(odata$x)), c(odata$mu + odata$sigma, rev(odata$mu - odata$sigma)),
  col = pallight["forest"], border = "transparent")
legend("topleft", expression(mu  %+-%  sigma), bty = "n")
@
\end{center}
\end{frame}

% \begin{frame}
% \frametitle{Distributional trees}
% \vspace*{-0.12cm}
% \begin{center} 
% DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
% 
% \vspace*{-0.21cm}
% \setkeys{Gin}{width=0.5\linewidth}
% <<plottree_dgp, fig=TRUE, echo=FALSE>>=
% paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
% plot(fig, inner_panel = node_inner_ext,
%      tp_args = list(FUN = identity, width = 18, fill = paltrees[c(1, 3)]), 
%      ip_args = list(fill = paltrees[c(2, 3)]),
%      drop_terminal = TRUE, tnex = 1.7)
% @
% \end{center}
% \end{frame}
